### 2014-09-10 src/paper1/p1_methods.tex
New siang command. Created custom style file.

Body tilt was controlled by a computer-controlled vestibular chair, which was configured to allow subject rotation in the roll axis. A digital position encoder measured roll position with an angular resolution of 0.04\textdegree. The subject's body was tightly fixated using a five-point seat belt and adjustable shoulder and hip supports. Velcro straps restrained both legs and feet, and a padded helmet firmly fixated the head in a natural upright position for looking straight ahead. Subject-specific seat adjustments ensured that the naso-occipital axis, midway between the eyes, coincided with the roll axis of the chair. Experiments took place in complete darkness.

@@

\newcommand{\mydegree}{\degree\ }
The SBT experiment served to obtain a psychometric measure of each subject's accuracy and precision of body-tilt perception at five body-tilt angles: upright (0\textdegree, \SBT{0} task) and 45\mydegree and 90\mydegree right side and left side down (\SBT{\textpm45} task and \SBT{\textpm90} task). Negative angles indicated left side down. We applied the method of constant stimuli, using a set of 10 equidistant body-tilt angles, centered on tentative estimates of the subject's 0\mydegree (SBT0), 45\mydegree (\SBT{45}), -45\mydegree (\SBT{-45}), 90\mydegree (\SBT{90}), and -90\mydegree (SBT-90) body-tilt percept. The latter were determined in a few pilot trials that also served to familiarize the subject with the task, without providing a reference of the five respective orientations to be tested. Relative to the test angle, we used test angle intervals of 3\textdegree, 4\textdegree, and 4\mydegree in the \SBT{0}, \SBT{\textpm45}, and \SBT{\textpm90} tasks, respectively. Body-tilt angles were tested 14 times in random order, yielding 140 responses for each psychometric curve.

To perform the psychophysical SBT experiments, two methodological problems had to be solved. The first relates to the number of experimental sessions that we could reasonably ask subjects to perform. We realized that returning the subject to upright for reorientation after each trial would require too large a number of experimental sessions. Our overriding concern was that starting each trial from upright would confound the \SBT{0} task in the sense that subjects could then simply notice the change in chair position. To prevent this, we always inserted a detour rotation before moving the subject to the test angle in a given trial. The detour, always to a tilt position clearly outside the psychometric test range, served to reset the subject's memory of the previous tilt position. These detour angles were chosen randomly from a range at 30-40\mydegree clockwise (CW) and counterclockwise (CCW) from the presumed threshold. As an illustration, Figure \ref{p1:fig2} shows how the subject was moved from one trial to the next in the course of an \SBT{90} experiment. Detour angles preceding each test angle were taken from the CW and the CCW detour range in equal proportions. An analysis of trial history effects indicated that detour angles did not affect the judgment in the subsequent trial ($p > 0.05$).

@@

    \caption{Tilt paradigm in \SBT{90} task. T1, T2, T3, Test angles at which the subject was prompted with a beep signal (*) to indicate whether body orientation was CW or CCW from the instructed reference orientation (i.e., 90\mydegree in this example). D1, D2, Detour angles randomly drawn from detour range (30 - 40\mydegree CW and CCW from center of test range). Rotations from detour (D) to test (T) angle were performed in a noisy fashion (see \nameref{p1:sec:methods}, \nameref{p1:sec:methods_sbt}).}
    \label{p1:fig2}
\end{wrapfigure}

Each experimental run started in upright position with the room lights on. After the lights were turned off, subjects were first rotated at a constant angular velocity of 30\textdegree/s to a random detour angle, outside of the test angle range, where they remained for 3 s. The chair then moved to a randomly chosen position within the test range with a very slow and noisy profile, defined by the sum of a ramp of 0.4 - 2\textdegree/s and Gaussian white noise (bandwidth, 0 - 0.7 Hz; RMS amplitude, 3.4\textdegree). Ramp speed was chosen such that the trajectory between detour angle and test angle was reached in 30 s (Fig. \ref{p1:fig2}). These precautions were taken to enforce independent absolute tilt judgments and to deter reliance on sensed changes in tilt position that had occurred since the previous trial. Three seconds after arrival at the test angle, a beep signal prompted the subject to indicate whether body orientation was CW or CCW from the instructed reference orientation (0\mydegree in the \SBT{0} task, \textpm45\mydegree in the \SBT{\textpm45}, or \textpm90\mydegree in the \SBT{\textpm90} task), using a toggle switch. The subject was then rotated at a constant velocity to a new randomly drawn detour angle, and the above procedure was repeated. Each run, comprising seven test angles, lasted \textapprox5 min, after which the subject was rotated back to upright, and room lights were turned on. Between runs, there was a 60 s rest interval before the next run started. Each SBT task was tested in separate sessions of \textapprox45 min each, thus amounting to a total of 15 sessions per subject (i.e., \textapprox11 h of recording time).

@@

The same subjects were also tested in a series of SVV experiments. Part of this dataset (four subjects) has been published previously as part of a larger dataset on visual verticality perception \cite{devrijer2009}. Data in the other three subjects were collected anew. Here we provide a brief summary of the paradigm. SVV was tested at nine roll-tilt angles, ranging from -120 to 120\mydegree at 30\mydegree intervals. A luminous line (angular subtend, 20\textdegree), polarized with a bright dot at one end, was mounted in front of the subject. The line's rotation axis coincided with the chair rotation axis. In each experimental run, the subject was rotated from upright to the chosen test angle at a constant angular velocity of 30\textdegree/s. After a 30 s waiting period that allowed canal effects to subside, a luminous line was briefly flashed (20 ms), and the subject indicated whether its orientation in space was CW or CCW from the perceived direction of gravity. The line orientation was selected randomly from a set of 11 line orientations. After all line orientations had been tested, the subject was rotated back to upright, and room lights were turned on. Positive and negative body-tilt angles were alternated regularly. As in the SBT experiment, we used the method of constant stimuli. The set of 11 line orientations was centered on a coarse estimate of the SVV threshold at each tilt angle. We used orientation intervals of 3\textdegree, except for upright, where intervals of 2\mydegree were taken. Each set of line orientations was tested in random order in 12 experimental runs, thus yielding a total of 132 responses for each psychometric curve. SVV data were collected in a total of five 45 min sessions per subject.

@@

CW tilt angles of the body and the luminous line were defined positive. We quantified performance, for each roll-tilt angle (5 in the SBT and 9 in the SVV) independently, by examining the proportion of CW responses as a function of body orientation (SBT) and the proportion of CCW responses as a function of line orientation with respect to the body (SVV). Psychometric data were quantified by fitting a cumulative Gaussian function (Fig. \ref{p1:fig3}):

@@

In the scheme (Fig. \ref{p1:fig1}), we use the following conventions: physical variables are denoted by a capital with a subscript indicating the frame of reference. For example, $H_S$ represents the physical orientation of the head in space. Sensory signals and their reference-frame-transformed counterparts are denoted by a hat symbol (\textasciicircum), as in $\hat{H}_{S}$, which represents the orientation of the head in space as measured by the head-in-space sensors. The optimal estimate of a variable, obtained by integration of all available information, is indicated by a tilde (\textasciitilde), as in $\tilde{H}_S$, representing the final head-in-space estimate.

@@

To obtain an estimate of the orientation of the body in space, the brain can use ``direct'' sensory information from body sensors ($\hat{B}_S$), such as tactile receptors in the skin or so-called graviceptors in the trunk \cite{mittelstaedt1997, mittelstaedt1998, vaitl2002}. Alternatively, an ``indirect'' pathway, involving a reference frame transformation, can also provide a body-in-space estimate. For this purpose, sensory head-in-space information, provided by the otoliths, must be combined with information about head-on-body orientation, provided by proprioceptive signals from the neck ($\hat{B}_{SI} = \hat{H}_S - \hat{H}_B$). Because the sensors are contaminated by noise, the direct and indirect signals can be represented as Gaussian probability distributions with mean values of $\hat{B}_S$ and $\hat{H}_S - \hat{H}_B$, and variance levels of $\sigma_{BS}^2$ and $\sigma_{HS}^2 + \sigma_{HB}^2$, respectively. Theoretically, as shown in Figure \ref{p1:fig1}, the brain could also use prior information about body-in-space orientation in the computation of the body-in-space estimate. The effect of including a prior on the SBT (centered on upright) would be a systematic error of underestimation at larger tilt angles. However, neither previous findings \cite{mittelstaedt1983, mast1996, jarchow1999, vanbeuzekom2001} nor the present results (Figs. \ref{p1:fig3}, \ref{p1:fig4}), showed such systematic errors across subjects. In modeling terms, this indicates a uniform (uninformative) prior, which corresponds to a weight of 0. Accordingly, a statistically optimal estimate of body-in-space orientation ($\tilde{B}_S$) is then given by the peak of the Gaussian distribution that results from the multiplication of the two distributions representing the direct and indirect sensory pathways. It follows that

@@

in which $w_{BD}$ and $w_{BI}$ (Fig. \ref{p1:fig1}) represent the respective weights of the direct and indirect pathways, which add up to 1 \cite{landy1995,jacobs1999,ernst2002,bays2007}. Note that the weight of each pathway depends on its reciprocal noise variance (also known as precision), so that precise signals have a stronger influence on the final estimate than noisy signals. Furthermore, because both sensory pathways are supposed to carry unbiased signals, the mean estimate of body in space in multiple trials, $\mu(\tilde{B}_S)$, will also be accurate.

@@

The scheme applies a similar sensory signal processing strategy to estimate the orientation of head in space, $\tilde{H}_S$, used in the SVV. A direct estimate of head-in-space orientation is provided by the head-in-space sensors ($\hat{H}_S$), and an indirect estimate is obtained by a reference frame transformation of the body-in-space signal ($\hat{B}_S$) by adding the head-on-body estimate ($\hat{H}_B$), provided by the neck sensors ($\hat{H}_{SI} = \hat{B}_S + \hat{H}_B$). Again, direct and indirect pathway signals are represented by two Gaussian probability distributions, with mean values of $\hat{H}_S$ and $\hat{B}_S + \hat{H}_B$, respectively, and corresponding variances of $\sigma^2_{HS}$ and $\sigma^2_{BS} + \sigma^2_{HB}$. In the computation of the head-in-space estimate, to account for systematic errors \cite{macneilage2007, devrijer2008}, it is further assumed that the brain uses prior knowledge about head-in-space orientation, which entails that small head-tilt angles are considered more probable than large tilts. Mathematically, the prior is represented by a Gaussian distribution that is centered at 0\mydegree head tilt ($H_{SP} = 0^\circ$) with a variance of $\sigma^2_{HSP}$. Note that, in our scheme, the head-in-space prior, which contributes to the SVV computations, does not affect the body-in-space estimate. Integration of the direct and indirect sensory pathways and prior knowledge is performed by multiplication of the three Gaussian distributions. The peak of the resulting posterior distribution represents the optimal estimate of head-in-space orientation ($\tilde{H}_S$), which is given by the following:

@@

In this equation, $w_{HD}$, $w_{HI}$, and $w_{HP}$ (which add up to one) represent the weights of the direct and indirect pathways and the prior, respectively, which are proportional to the relative precision of the sensory signals and the width of the prior. Equation \ref{p1:eqn6} would result in an accurate estimate of $\tilde{H}_S$, if all three pathways were accurate by themselves. However, because the prior is centered on zero ($H_{SP} = 0\degree$), it introduces more and more bias toward upright, as head tilt increases further. Thus, optimization in terms of variance has a downside by causing underestimation of the actual head tilt. The amount of underestimation depends on the width of the prior and the reliability of the sensory inputs.

The variance in the head-in-space estimates, measured across many trials, $\sigma^2(\tilde{H}_S)$, can be derived directly from Equation \ref{p1:eqn6} by applying the rules of error propagation (see \nameref{p1:sec:appendix} for complete derivation). From these calculations, it follows that

@@

in which the variance contributions of the direct and indirect pathways are represented by their squared weights. Although it does not appear explicitly in Equation \ref{p1:eqn10}, it is important to notice that the prior has a noise-reducing effect by downscaling the sensory-related weighting terms ($w_{HD}$ and $w_{HI}$). The narrower the prior, the larger its relative weight ($w_{HP}$) and the smaller the sensory weights, because $w_{HD} + w_{HI} + w_{HP} = 1$. Thus, the effect of the head-in-space prior is twofold: it reduces the variance, but as noticed above, this occurs at the cost of a bias in the final estimate of head-in-space orientation, which becomes pronounced at large tilts (see \nameref{p1:sec:appendix} for further details).

@@

To compute the SVV, the brain not only requires an estimate of head orientation in space ($\tilde{H}_S$), but also needs estimates of eye-in-head orientation ($\tilde{E}_H$) and retinal line orientation ($\tilde{L}_E$). Together, these signals determine the orientation of a visual line in space ($\tilde{L}_S$) according to $\tilde{L}_S = \tilde{H}_S + \tilde{E}_H + \tilde{L}_E$. The systematic error in the SVV experiment (${\Delta}SVV$) corresponds to the error in $\tilde{L}_S$ and is thus given by ${\Delta}SVV = {\Delta}H_S + {\Delta}E_H + {\Delta}L_E$, in which $\Delta$ denotes the bias in each estimate. For simplicity, we assumed that the visual signal representing retinal line orientation is accurate, so that ${\Delta}L_E = 0\degree$. As explained in a previous study \cite{devrijer2009}, underestimation of eye torsion causes errors in the eye-in-head estimate (${\Delta}E_H$), which can be represented by ${\Delta}E_H = -A_{OCR} \cdot \sin(\hat{H}_S$, where parameter $A_{OCR}$ denotes uncompensated ocular counterroll. Finally, the error in the head-in-space estimate is obtained by subtracting $\tilde{H}_S$ (see Eq. \ref{p1:eqn6}) from the actual head tilt $H_S$, which ultimately leads to the following relation for the mean SVV error, $\mu({\Delta}SVV)$, in multiple trials:

@@

In Equation \ref{p1:eqn12}, the influence of the prior works through the weight factors $w_{HD}$ and $w_{HI}$. Because these weights do not add up to 1 (see above, $w_{HD} + w_{HI} = 1 - w_{HP} < 1$), the result is a systematic error in the head-in-space estimate, which becomes more pronounced at large tilt. The noise level in the eye-in-head and line-on-eye estimates is probably relatively small compared with the noise in the head-in-space estimate considering results from Vandenbussche et al. \citeyear{vandenbussche1986}, who reported just-noticeable difference levels for orientation discrimination of \textless1\textdegree. Given this low value, SVV variance is determined mainly by the variance in the latter estimate, so that $\sigma^2({\Delta}SVV) \sim \sigma^2(\tilde{H}_S)$.

@@

The SBT and SVV measurements to test the model proposed in Figure 1 have yielded solutions for the noise properties of the involved sensor systems. To validate the model structure and the noise predictions that were obtained, we also devised an experiment that independently measured the noise in the neck sensors (head-on-body sensors), in a psychometric fashion. In this experiment, subjects were lying on a bed, in supine position, with their head fixed on a rotating platform. The platform was constructed such that it could passively rotate the head relative to the body, in the roll plane, while accounting for the shifting rotation axis in the neck vertebrae. The rotation of the platform was computer controlled, keeping the speed below $0.2{\degree}/s$, which is far below detection threshold of the canals ($>0.5{\degree}/s$) (Benson et al., 1989). In the supine condition, there is no gravity modulation of the otolith signal, so we excluded the contribution of the vestibular system in detecting head-on-body orientation and were only probing the role of the neck afferents. We applied the method of constant stimuli, using a set of 11 head angles relative to body midline. Test angles ranged from $-6\degree$ to $6\degree$.

In complete darkness, subjects were first rotated at a constant angular velocity of =15∞/s to a random detour angle similar to the idea shown in Figure 2. The head then moved to a randomly chosen position within the test range with a very slow speed ($<0.2{\degree}/s$) such that the test angle was reached within 20s. Meanwhile, auditory white noise was presented to the subjects through earphones to mask any auditory cues generated by the moving platform. After arrival at the test angle, the auditory noise was interrupted, signaling the subject to indicate whether head-on-body orientation was CW or CCW relative to the body midline, using a toggle switch. The subject was then rotated at a constant velocity to a new randomly chosen detour angle, and the above procedure was repeated. Each test angle was repeated 10 times, yielding a total number of 110 responses in each subject. Psychometric data were quantified by fitting a cumulative Gaussian function (see above, Eq. 1). The width of the curve, $\sigma^2$, inversely related to precision, serves as an independent measure of the subject's variability of the head-on-body estimate and was compared with the model prediction.

>>>>

Body tilt was controlled by a computer-controlled vestibular chair, which was configured to allow subject rotation in the roll axis. A digital position encoder measured roll position with an angular resolution of \siang{0.04}. The subject's body was tightly fixated using a five-point seat belt and adjustable shoulder and hip supports. Velcro straps restrained both legs and feet, and a padded helmet firmly fixated the head in a natural upright position for looking straight ahead. Subject-specific seat adjustments ensured that the naso-occipital axis, midway between the eyes, coincided with the roll axis of the chair. Experiments took place in complete darkness.

@@

The SBT experiment served to obtain a psychometric measure of each subject's accuracy and precision of body-tilt perception at five body-tilt angles: upright (\siang{0}, \SBT{0} task) and \siang{45} and \siang{90} right side and left side down (\SBT{\textpm45} task and \SBT{\textpm90} task). Negative angles indicated left side down. We applied the method of constant stimuli, using a set of 10 equidistant body-tilt angles, centered on tentative estimates of the subject's \siang{0} (SBT0), \siang{45} (\SBT{45}), \siang{-45} (\SBT{-45}), \siang{90} (\SBT{90}), and \siang{-90} (SBT-90) body-tilt percept. The latter were determined in a few pilot trials that also served to familiarize the subject with the task, without providing a reference of the five respective orientations to be tested. Relative to the test angle, we used test angle intervals of \siang{3}, \siang{4}, and \siang{4} in the \SBT{0}, \SBT{\textpm45}, and \SBT{\textpm90} tasks, respectively. Body-tilt angles were tested 14 times in random order, yielding 140 responses for each psychometric curve.

To perform the psychophysical SBT experiments, two methodological problems had to be solved. The first relates to the number of experimental sessions that we could reasonably ask subjects to perform. We realized that returning the subject to upright for reorientation after each trial would require too large a number of experimental sessions. Our overriding concern was that starting each trial from upright would confound the \SBT{0} task in the sense that subjects could then simply notice the change in chair position. To prevent this, we always inserted a detour rotation before moving the subject to the test angle in a given trial. The detour, always to a tilt position clearly outside the psychometric test range, served to reset the subject's memory of the previous tilt position. These detour angles were chosen randomly from a range at \siang{30-40} clockwise (CW) and counterclockwise (CCW) from the presumed threshold. As an illustration, \figref{p1:fig2} shows how the subject was moved from one trial to the next in the course of an \SBT{90} experiment. Detour angles preceding each test angle were taken from the CW and the CCW detour range in equal proportions. An analysis of trial history effects indicated that detour angles did not affect the judgment in the subsequent trial ($p > 0.05$).

@@

    \caption{Tilt paradigm in \SBT{90} task. T1, T2, T3, Test angles at which the subject was prompted with a beep signal (*) to indicate whether body orientation was CW or CCW from the instructed reference orientation (i.e., \siang{90} in this example). D1, D2, Detour angles randomly drawn from detour range (\siang{30-40} CW and CCW from center of test range). Rotations from detour (D) to test (T) angle were performed in a noisy fashion (see \nameref{p1:sec:methods}, \nameref{p1:sec:methods_sbt}).}
    \label{p1:fig2}
\end{wrapfigure}

Each experimental run started in upright position with the room lights on. After the lights were turned off, subjects were first rotated at a constant angular velocity of \siang[/s]{30} to a random detour angle, outside of the test angle range, where they remained for 3 s. The chair then moved to a randomly chosen position within the test range with a very slow and noisy profile, defined by the sum of a ramp of \siang[/s]{0.4 - 2} and Gaussian white noise (bandwidth, 0 - 0.7 Hz; RMS amplitude, \siang{3.4}). Ramp speed was chosen such that the trajectory between detour angle and test angle was reached in 30 \si{\second} (\figref{p1:fig2}). These precautions were taken to enforce independent absolute tilt judgments and to deter reliance on sensed changes in tilt position that had occurred since the previous trial. Three seconds after arrival at the test angle, a beep signal prompted the subject to indicate whether body orientation was CW or CCW from the instructed reference orientation (\siang{0} in the \SBT{0} task, \siang{\textpm45} in the \SBT{\textpm45}, or \siang{\textpm90} in the \SBT{\textpm90} task), using a toggle switch. The subject was then rotated at a constant velocity to a new randomly drawn detour angle, and the above procedure was repeated. Each run, comprising seven test angles, lasted \textapprox5 min, after which the subject was rotated back to upright, and room lights were turned on. Between runs, there was a 60 s rest interval before the next run started. Each SBT task was tested in separate sessions of \textapprox45 min each, thus amounting to a total of 15 sessions per subject (i.e., \textapprox11 h of recording time).

@@

The same subjects were also tested in a series of SVV experiments. Part of this dataset (four subjects) has been published previously as part of a larger dataset on visual verticality perception \cite{devrijer2009}. Data in the other three subjects were collected anew. Here we provide a brief summary of the paradigm. SVV was tested at nine roll-tilt angles, ranging from -120 to \siang{120} at \siang{30} intervals. A luminous line (angular subtend, \siang{20}), polarized with a bright dot at one end, was mounted in front of the subject. The line's rotation axis coincided with the chair rotation axis. In each experimental run, the subject was rotated from upright to the chosen test angle at a constant angular velocity of \siang[/s]{30}. After a 30 \si{\second} waiting period that allowed canal effects to subside, a luminous line was briefly flashed (20 \si{\milli\second}), and the subject indicated whether its orientation in space was CW or CCW from the perceived direction of gravity. The line orientation was selected randomly from a set of 11 line orientations. After all line orientations had been tested, the subject was rotated back to upright, and room lights were turned on. Positive and negative body-tilt angles were alternated regularly. As in the SBT experiment, we used the method of constant stimuli. The set of 11 line orientations was centered on a coarse estimate of the SVV threshold at each tilt angle. We used orientation intervals of \siang{3}, except for upright, where intervals of \siang{2} were taken. Each set of line orientations was tested in random order in 12 experimental runs, thus yielding a total of 132 responses for each psychometric curve. SVV data were collected in a total of five 45 min sessions per subject.

@@

CW tilt angles of the body and the luminous line were defined positive. We quantified performance, for each roll-tilt angle (5 in the SBT and 9 in the SVV) independently, by examining the proportion of CW responses as a function of body orientation (SBT) and the proportion of CCW responses as a function of line orientation with respect to the body (SVV). Psychometric data were quantified by fitting a cumulative Gaussian function (\figref{p1:fig3}):

@@

In the scheme (\figref{p1:fig1}), we use the following conventions: physical variables are denoted by a capital with a subscript indicating the frame of reference. For example, $H_S$ represents the physical orientation of the head in space. Sensory signals and their reference-frame-transformed counterparts are denoted by a hat symbol (\textasciicircum), as in $\hat{H}_{S}$, which represents the orientation of the head in space as measured by the head-in-space sensors. The optimal estimate of a variable, obtained by integration of all available information, is indicated by a tilde (\textasciitilde), as in $\tilde{H}_S$, representing the final head-in-space estimate.

@@

To obtain an estimate of the orientation of the body in space, the brain can use ``direct'' sensory information from body sensors ($\hat{B}_S$), such as tactile receptors in the skin or so-called graviceptors in the trunk \cite{mittelstaedt1997, mittelstaedt1998, vaitl2002}. Alternatively, an ``indirect'' pathway, involving a reference frame transformation, can also provide a body-in-space estimate. For this purpose, sensory head-in-space information, provided by the otoliths, must be combined with information about head-on-body orientation, provided by proprioceptive signals from the neck ($\hat{B}_{SI} = \hat{H}_S - \hat{H}_B$). Because the sensors are contaminated by noise, the direct and indirect signals can be represented as Gaussian probability distributions with mean values of $\hat{B}_S$ and $\hat{H}_S - \hat{H}_B$, and variance levels of $\sigma_{BS}^2$ and $\sigma_{HS}^2 + \sigma_{HB}^2$, respectively. Theoretically, as shown in \figref{p1:fig1}, the brain could also use prior information about body-in-space orientation in the computation of the body-in-space estimate. The effect of including a prior on the SBT (centered on upright) would be a systematic error of underestimation at larger tilt angles. However, neither previous findings \cite{mittelstaedt1983, mast1996, jarchow1999, vanbeuzekom2001} nor the present results (\figref[ and ]{p1:fig3,p1:fig4}), showed such systematic errors across subjects. In modeling terms, this indicates a uniform (uninformative) prior, which corresponds to a weight of 0. Accordingly, a statistically optimal estimate of body-in-space orientation ($\tilde{B}_S$) is then given by the peak of the Gaussian distribution that results from the multiplication of the two distributions representing the direct and indirect sensory pathways. It follows that

@@

in which $w_{BD}$ and $w_{BI}$ (\figref{p1:fig1}) represent the respective weights of the direct and indirect pathways, which add up to 1 \cite{landy1995,jacobs1999,ernst2002,bays2007}. Note that the weight of each pathway depends on its reciprocal noise variance (also known as precision), so that precise signals have a stronger influence on the final estimate than noisy signals. Furthermore, because both sensory pathways are supposed to carry unbiased signals, the mean estimate of body in space in multiple trials, $\mu(\tilde{B}_S)$, will also be accurate.

@@

The scheme applies a similar sensory signal processing strategy to estimate the orientation of head in space, $\tilde{H}_S$, used in the SVV. A direct estimate of head-in-space orientation is provided by the head-in-space sensors ($\hat{H}_S$), and an indirect estimate is obtained by a reference frame transformation of the body-in-space signal ($\hat{B}_S$) by adding the head-on-body estimate ($\hat{H}_B$), provided by the neck sensors ($\hat{H}_{SI} = \hat{B}_S + \hat{H}_B$). Again, direct and indirect pathway signals are represented by two Gaussian probability distributions, with mean values of $\hat{H}_S$ and $\hat{B}_S + \hat{H}_B$, respectively, and corresponding variances of $\sigma^2_{HS}$ and $\sigma^2_{BS} + \sigma^2_{HB}$. In the computation of the head-in-space estimate, to account for systematic errors \cite{macneilage2007, devrijer2008}, it is further assumed that the brain uses prior knowledge about head-in-space orientation, which entails that small head-tilt angles are considered more probable than large tilts. Mathematically, the prior is represented by a Gaussian distribution that is centered at \siang{0} head tilt ($H_{SP} = 0^\circ$) with a variance of $\sigma^2_{HSP}$. Note that, in our scheme, the head-in-space prior, which contributes to the SVV computations, does not affect the body-in-space estimate. Integration of the direct and indirect sensory pathways and prior knowledge is performed by multiplication of the three Gaussian distributions. The peak of the resulting posterior distribution represents the optimal estimate of head-in-space orientation ($\tilde{H}_S$), which is given by the following:

@@

In this equation, $w_{HD}$, $w_{HI}$, and $w_{HP}$ (which add up to one) represent the weights of the direct and indirect pathways and the prior, respectively, which are proportional to the relative precision of the sensory signals and the width of the prior. \eqnref{p1:eqn6} would result in an accurate estimate of $\tilde{H}_S$, if all three pathways were accurate by themselves. However, because the prior is centered on zero ($H_{SP} = 0\degree$), it introduces more and more bias toward upright, as head tilt increases further. Thus, optimization in terms of variance has a downside by causing underestimation of the actual head tilt. The amount of underestimation depends on the width of the prior and the reliability of the sensory inputs.

The variance in the head-in-space estimates, measured across many trials, $\sigma^2(\tilde{H}_S)$, can be derived directly from \eqnref{p1:eqn6} by applying the rules of error propagation (see \nameref{p1:sec:appendix} for complete derivation). From these calculations, it follows that

@@

in which the variance contributions of the direct and indirect pathways are represented by their squared weights. Although it does not appear explicitly in \eqnref{p1:eqn10}, it is important to notice that the prior has a noise-reducing effect by downscaling the sensory-related weighting terms ($w_{HD}$ and $w_{HI}$). The narrower the prior, the larger its relative weight ($w_{HP}$) and the smaller the sensory weights, because $w_{HD} + w_{HI} + w_{HP} = 1$. Thus, the effect of the head-in-space prior is twofold: it reduces the variance, but as noticed above, this occurs at the cost of a bias in the final estimate of head-in-space orientation, which becomes pronounced at large tilts (see \nameref{p1:sec:appendix} for further details).

@@

To compute the SVV, the brain not only requires an estimate of head orientation in space ($\tilde{H}_S$), but also needs estimates of eye-in-head orientation ($\tilde{E}_H$) and retinal line orientation ($\tilde{L}_E$). Together, these signals determine the orientation of a visual line in space ($\tilde{L}_S$) according to $\tilde{L}_S = \tilde{H}_S + \tilde{E}_H + \tilde{L}_E$. The systematic error in the SVV experiment (${\Delta}SVV$) corresponds to the error in $\tilde{L}_S$ and is thus given by ${\Delta}SVV = {\Delta}H_S + {\Delta}E_H + {\Delta}L_E$, in which $\Delta$ denotes the bias in each estimate. For simplicity, we assumed that the visual signal representing retinal line orientation is accurate, so that ${\Delta}L_E = 0\degree$. As explained in a previous study \cite{devrijer2009}, underestimation of eye torsion causes errors in the eye-in-head estimate (${\Delta}E_H$), which can be represented by ${\Delta}E_H = -A_{OCR} \cdot \sin(\hat{H}_S$, where parameter $A_{OCR}$ denotes uncompensated ocular counterroll. Finally, the error in the head-in-space estimate is obtained by subtracting $\tilde{H}_S$ (see \eqnref{p1:eqn6}) from the actual head tilt $H_S$, which ultimately leads to the following relation for the mean SVV error, $\mu({\Delta}SVV)$, in multiple trials:

@@

In \eqnref{p1:eqn12}, the influence of the prior works through the weight factors $w_{HD}$ and $w_{HI}$. Because these weights do not add up to 1 (see above, $w_{HD} + w_{HI} = 1 - w_{HP} < 1$), the result is a systematic error in the head-in-space estimate, which becomes more pronounced at large tilt. The noise level in the eye-in-head and line-on-eye estimates is probably relatively small compared with the noise in the head-in-space estimate considering results from Vandenbussche et al. \citeyear{vandenbussche1986}, who reported just-noticeable difference levels for orientation discrimination of \siang{\textless1}. Given this low value, SVV variance is determined mainly by the variance in the latter estimate, so that $\sigma^2({\Delta}SVV) \sim \sigma^2(\tilde{H}_S)$.

@@

The SBT and SVV measurements to test the model proposed in \figref{p1:fig1} have yielded solutions for the noise properties of the involved sensor systems. To validate the model structure and the noise predictions that were obtained, we also devised an experiment that independently measured the noise in the neck sensors (head-on-body sensors), in a psychometric fashion. In this experiment, subjects were lying on a bed, in supine position, with their head fixed on a rotating platform. The platform was constructed such that it could passively rotate the head relative to the body, in the roll plane, while accounting for the shifting rotation axis in the neck vertebrae. The rotation of the platform was computer controlled, keeping the speed below \siang[/s]{0.2}, which is far below detection threshold of the canals \siang[/s]{$>0.5$}) \cite{benson1989}. In the supine condition, there is no gravity modulation of the otolith signal, so we excluded the contribution of the vestibular system in detecting head-on-body orientation and were only probing the role of the neck afferents. We applied the method of constant stimuli, using a set of 11 head angles relative to body midline. Test angles ranged from $-6\degree$ to $6\degree$.

In complete darkness, subjects were first rotated at a constant angular velocity of \siang[/s]{$\leq15$} to a random detour angle similar to the idea shown in \figref{p1:fig2}. The head then moved to a randomly chosen position within the test range with a very slow speed (\siang[/s]{$<0.2$}) such that the test angle was reached within 20 \si{\second}. Meanwhile, auditory white noise was presented to the subjects through earphones to mask any auditory cues generated by the moving platform. After arrival at the test angle, the auditory noise was interrupted, signaling the subject to indicate whether head-on-body orientation was CW or CCW relative to the body midline, using a toggle switch. The subject was then rotated at a constant velocity to a new randomly chosen detour angle, and the above procedure was repeated. Each test angle was repeated 10 times, yielding a total number of 110 responses in each subject. Psychometric data were quantified by fitting a cumulative Gaussian function (see above, \eqnref{p1:eqn1}). The width of the curve, $\sigma^2$, inversely related to precision, serves as an independent measure of the subject's variability of the head-on-body estimate and was compared with the model prediction.
